---
title: "README"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## PowerTOST

The package contains functions to calculate power and sample size for various study designs used in bioequivalence studies.

### Supported designs

```{r, echo = FALSE}
designs <- PowerTOST:::known.designs()
print(designs[, c(2, 9, 3)], row.names = FALSE)
```

### Purpose

For various methods power can be *calculated* based on

  - nominal *α*, coefficient of variation (*CV*), deviation of test from reference (*θ*<sub>0</sub>), accteptance limits {*θ*<sub>1</sub>, *θ*<sub>2</sub>}, sample size (*n*), and design.

For all methods the sample size can be *estimated* based on

  - nominal *α*, coefficient of variation (*CV*), deviation of test from reference(*θ*<sub>0</sub>), accteptance limits {*θ*<sub>1</sub>, *θ*<sub>2</sub>}, target power, and design.

### Supported methods
#### Power and sample size
  - Average bioequivalence (with arbitrary *fixed* limits).
  - Reference-scaled bioequivalence (EMA: ABEL, FDA: RSABE for HVD(P)s and NTIDs).
  - Iteratively adjust *α* to control the type I error in ABEL and RSABE.
  - Two simulataneous TOST procedures.
  - Dose proportionality using the power model.
  - Non-inferiority *t*-test.
  - Ratio of two means with normally distributed data on the original scale based on Fieller’s (‘fiducial’) confidence interval.
  - ‘Expected’ power in case of uncertain (estimated) variability  and/or uncertain *θ*<sub>0</sub>.

#### Helpers
  - Calculate *CV* from *MSE* or *SE* (and vice versa).
  - Calculate *CV* from given confidence interval.
  - Confidence interval of *CV*.
  - Pool *CV* from several studies.
  - Confidence interval for given *α*, *CV*, point estimate, sample size, and design.
  - *p*-values of the TOST procedure.
  - Analysis tool for exploration/visualization of the impact of expected values (*CV*, *θ*<sub>0</sub>, reduced sample size due to drop-outs) on power of BE decision.

### Examples

  1. Power of a paralled design (group sizes 75 and 70), total *CV* 0.36, *θ*<sub>0</sub> 0.925. The defaults for *α* (0.05) and {*θ*<sub>1</sub>, *θ*<sub>2</sub>} (0.80, 1.25) are automatically employed.
```{r}
PowerTOST:::power.TOST(CV = 0.36, theta0 = 0.93,
                       n = c(75, 70), design = "parallel")
```
  2. Sample size for a 2×2×2 crossover study, assumed intra-subject *CV* 0.20, and desired power 0.80. The defaults for *α* (0.05), *θ*<sub>0</sub> (0.95), {*θ*<sub>1</sub>, *θ*<sub>2</sub>} (0.80, 1.25), targetpower (0.80), and design ("2x2") are automatically employed.
```{r}
PowerTOST:::sampleN.TOST(CV = 0.20)
```
  3. Sample size for a four-period replicate study intended for the EMA’s average bioequivalence with expanding limits (ABEL), assumed heteroscedasticity (*CV<sub>wT</sub>* 0.40, *CV<sub>wR</sub>* 0.50). The defaults for *α* (0.05), *θ*<sub>0</sub> (0.90 for highly variable drugs / drug products), and targetpower (0.80) are automatically employed. The expanded limits {*θ*<sub>1</sub>, *θ*<sub>2</sub>} are calculated based on *CV<sub>wR</sub>*.
```{r}
PowerTOST:::sampleN.scABEL(CV = c(0.40, 0.50), design = "2x2x4")
```
### Installation
```{r}
package <- "PowerTOST"
inst    <- package %in% installed.packages()
if (length(package[!inst]) > 0) install.packages(package[!inst])
```
