---
title: "PowerTOST"
output:
  github_document:
    toc: true
    toc_depth: 3
---
<!-- README.md is generated from README.Rmd. Please edit that file -->
[![License: GPL v2](https://img.shields.io/badge/License-GPLv2-blue.svg)](https://www.gnu.org/licenses/gpl-2.0) [![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
![active](https://www.repostatus.org/badges/latest/active.svg) ![on CRAN](https://www.r-pkg.org/badges/version-ago/PowerTOST) [![cran checks](https://cranchecks.info/badges/worst/PowerTOST)](https://cran.r-project.org/web/checks/check_results_PowerTOST.html) [![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/grand-total/PowerTOST?color=blue)](https://r-pkg.org/pkg/PowerTOST) [![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/last-month/PowerTOST?color=green)](https://r-pkg.org/pkg/PowerTOST)
```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```
The package contains functions to calculate power and estimate sample size for various study designs used in (not only bio-) equivalence studies.  
Published on CRAN `r packageDate("PowerTOST", date.fields = "Date/Publication")` (version `r packageVersion("PowerTOST")` built `r packageDate("PowerTOST", date.fields = "Built")` with R `r substr(packageDescription("PowerTOST", fields = "Built"), 3, 7)`).

## Supported Designs

```{r, echo = FALSE}
library(PowerTOST)
designs <- known.designs()
print(designs[, c(2, 9, 3)], row.names = FALSE)
```

Codes of designs follow this pattern: `treatments x sequences x periods`.

Although some replicate designs are more ‘popular’ than others, sample size estimations are valid for *all* of the following designs:

|design|type|sequences|periods|
|:----:|:----:|---------|:-:|
|`2x2x4`|full|2 `TRTR|RTRT`|4|
|`2x2x4`|full|2 `TRRT|RTTR`|4|
|`2x2x4`|full|2 `TTRR|RRTT`|4|
|`2x2x3`|full|2 `TRT|RTR`|3|
|`2x2x3`|full|2 `TRR|RTT`|3|
|`2x3x3`|partial|3 `TRR|RTR|RRT`|3|

Whilst \"2x4x4\" four period full replicate designs with *four* sequences (TRTR|RTRT|TRRT|RTTR *or* TRRT|RTTR|TTRR|RRTT) are supported, they should be avoided due to confounded effects.

<small>[TOC ↩](#powertost)</small>

## Purpose

For various methods power can be *calculated* based on

  - nominal *α*, coefficient of variation (*CV*), deviation of test from reference (*θ*~0~), acceptance limits {*θ*~1~, *θ*~2~}, sample size (*n*), and design.

For all methods the sample size can be *estimated* based on

  - nominal *α*, coefficient of variation (*CV*), deviation of test from reference (*θ*~0~), acceptance limits {*θ*~1~, *θ*~2~}, target (*i.e.*, desired) power, and design.

<small>[TOC ↩](#powertost)</small>

## Supported
### Power and Sample Size
Power covers balanced as well as unbalanced sequences in crossover or replicate designs and equal/unequal group sizes in two-group parallel designs. Sample sizes are always rounded up to achieve balanced sequences or equal group sizes.

  - Average Bioequivalence (with arbitrary *fixed* limits).
  - Two simultaneous <span title="Two One-Sided Tests">TOST</span> procedures.
  - Non-inferiority *t*-test.
  - Ratio of two means with normally distributed data on the original scale based on Fieller’s (‘fiducial’) confidence interval.
  - ‘Expected’ power in case of uncertain (estimated) variability and/or uncertain *θ*~0~.
  - Reference-scaled bioequivalence based on simulations.  
    - <span title="European Medicines Agency">EMA</span>: Average Bioequivalence with Expanding Limits (ABEL).  
    - U.S. <span title="Food and Drug Administration">FDA</span>, China <span title="Centre for Drug Evaluation">CDE</span>: Reference-scaled Average Bioequivalence (RSABE) for Highly Variable Drugs / Drug Products and Narrow Therapeutic Index Drugs (NTIDs). In China the former is required  and the latter currently under discussion.
  - Iteratively adjust *α* to control the type I error in ABEL and RSABE.
  - U.S. <span title="Food and Drug Administration">FDA</span>: <span title="Average Bioequivalence">ABE</span> for highly variable <span title="Narrow Therapeutic Index Drugs">NTIDs</span> by simulations.
  - Dose-Proportionality using the power model.

<small>[TOC ↩](#powertost)</small>

### Methods
  - Exact
    - Owen’s Q.
    - Direct integration of the bivariate non-central *t*-distribution.
  - Approximations
    - Non-central *t*-distribution.
    - ‘Shifted’ central *t*-distribution.

<small>[TOC ↩](#powertost)</small>

### Helpers
  - Calculate *CV* from *MSE* or *SE* (and vice versa).
  - Calculate *CV* from given confidence interval.
  - Calculate *CV*~wR~ from the upper expanded limit of an ABEL study.
  - Confidence interval of *CV*.
  - Pool *CV* from several studies.
  - Confidence interval for given *α*, *CV*, point estimate, sample size, and design.
  - Calculate *CV*~wT~ and *CV*~wR~ from a (pooled) *CV*~w~ assuming a ratio of intra-subject variances.
  - *p*-values of the <span title="Two One-Sided Tests">TOST</span> procedure.
  - Analysis tool for exploration/visualization of the impact of expected values (*CV*, *θ*~0~, reduced sample size due to dropouts) on power of BE decision.
  - Construct design matrices of incomplete block designs.

<small>[TOC ↩](#powertost)</small>

## Defaults {#defaults}
  * *α* 0.05, {*θ*~1~, *θ*~2~} (0.80, 1.25). Details of the sample size search (and the regulatory settings in reference-scaled average bioequivalence) are shown in the console.
  * Note: In all functions values have to be given as ratios, not in percent.  

### Average Bioequivalence  
*θ*~0~ 0.95, target power 0.80, design \"2x2\" (TR|RT), exact method (Owen’s Q).

### Reference-Scaled Average Bioequivalence  
*α* 0.05, point estimate constraint (0.80, 1.25), homoscedasticity (*CV*~wT~ = *CV*~wR~),  scaling is based on *CV*~wR~, target power 0.80, design \"2x3x3\" (TRR|RTR|RRT), approximation by the non-central *t*-distribution, 100,000 simulations.

  - <span title="European Medicines Agency">EMA</span>, <span title="World Health Organization">WHO</span>, Health Canada, and many other jurisdictions: Average Bioequivalence with Expanding Limits (ABEL).
  - U.S. <span title="Food and Drug Administration">FDA</span>, China <span title="Centre for Drug Evaluation">CDE</span>: RSABE.

#### Highly Variable Drugs / Drug Products  
*θ*~0~ 0.90 as recommended by [Tóthfalusi and Endrényi](https://ejournals.library.ualberta.ca/index.php/JPPS/article/download/11612/9489) (2011).

###### EMA
Regulatory constant `0.76`, upper cap of scaling at *CV*~wR~ 50\%, evaluation by <span title="Analysis of Variance">ANOVA</span>.

###### Health Canada
Regulatory constant `0.76`, upper cap of scaling at *CV*~wR~ ~57.4\%, evaluation by intra-subject contrasts.

###### FDA
Regulatory constant `log(1.25)/0.25`, linearized scaled <span title="Average Bioequivalence">ABE</span> (Howe’s approximation).

#### Narrow Therapeutic Index Drugs (FDA)
*θ*~0~ 0.975, regulatory constant `log(1.11111)/0.1`, upper cap of scaling at *CV*~wR~ ~21.4\%, design \"2x2x4\" (TRTR|RTRT), linearized scaled <span title="Average Bioequivalence">ABE</span> (Howe’s approximation), upper limit of the confidence interval of *s*~wT~/*s*~wR~ ≤2.5.

### Dose-Proportionality
*β*~0~ (slope) `1+log(0.95)/log(rd)` where `rd` is the ratio of the highest and lowest dose, target power 0.80, crossover design, details of the sample size search suppressed.

### Power Analysis
Minimum acceptable power 0.70. *θ*~0~, design, conditions, and sample size method depend on defaults of the respective approaches (ABE, ABEL, RSABE, NTID, HVNTID).

<small>[TOC ↩](#powertost)</small>

## Examples
Before running the examples attach the library.

```{r attach_library}
library(PowerTOST)
```
If not noted otherwise, the functions’ [defaults](#defaults) are employed.

### Parallel Design
Power for total *CV* 0.35 (35\%), group sizes 52 and 49, design \"parallel\".
```{r parallel}
power.TOST(CV = 0.35, n = c(52, 49), design = "parallel")
```

### Crossover Design
Sample size for assumed within- (intra-) subject *CV* 0.20 (20\%).
```{r crossover_ABE1}
sampleN.TOST(CV = 0.20)
```

Sample size for assumed within- (intra-) subject *CV* 0.40 (40\%), *θ*~0~ 0.90, four period full replicate "2x2x4" study. Wider acceptance range for *C*~max~ (Gulf Cooperation Council, South Africa).
```{r crossover_ABE2}
sampleN.TOST(CV = 0.40, theta0 = 0.90, theta1 = 0.75, design = "2x2x4")
```

<small>[TOC ↩](#powertost)</small>

Sample size for assumed within- (intra-) subject *CV* 0.125 (12.5\%), *θ*~0~ 0.975. Acceptance range for <span title="Narrow Therapeutic Index Drugs">NTIDs</span> (most jurisdictions).
```{r crossover_ABE3}
sampleN.TOST(CV = 0.125, theta0 = 0.975, theta1 = 0.90)
```

<small>[TOC ↩](#powertost)</small>

Sample size for equivalence of the ratio of two means with normality on the original scale based on [Fieller’s (‘fiducial’) confidence interval](https://dx.doi.org/10.1111/j.2517-6161.1954.tb00159.x). Within- (intra-) subject *CV*~w~ 0.20 (20\%), between- (inter-) subject *CV*~b~ 0.40 (40\%).  
Note the default *α* 0.025 (95\% CI) of this function because it is intended for studies with clinical endpoints.
```{r crossover_fieller}
sampleN.RatioF(CV = 0.20, CVb = 0.40)
```

<small>[TOC ↩](#powertost)</small>

### Replicate Designs

#### ABE
Sample size for assumed within- (intra-) subject *CV* 0.45 (45\%), *θ*~0~ 0.90, three period full replicate study \"2x2x3\" (TRT|RTR *or* TRR|RTT).
```{r replicate_ABE1}
sampleN.TOST(CV = 0.45, theta0 = 0.90, design = "2x2x3")
```
Note that the conventional model assumes homoscedasticity (equal variances of treatments). For heteroscedasticity we can ‘switch off’ all conditions of one of the methods for reference-scaled <span title="Average Bioequivalence">ABE</span>. We assume a σ^2^ ratio of ⅔ (*i.e.*, the test has a lower variability than the reference). Only relevant columns of the data.frame shown.
```{r replicate_ABE2}
reg <- reg_const("USER", r_const = NA, CVswitch = Inf,
                 CVcap = Inf, pe_constr = FALSE)
CV  <- CVp2CV(CV = 0.45, ratio = 2/3)
res <- sampleN.scABEL(CV=CV, design = "2x2x3", regulator = reg,
                      details = FALSE, print = FALSE)
print(res[c(3:4, 8:9)], digits = 5, row.names = FALSE)
```
Similar sample size because the pooled *CV* is still 0.45.

<small>[TOC ↩](#powertost)</small>

#### ABEL
Sample size assuming homoscedasticity (*CV*~wT~ = *CV*~wR~ = 0.45).
```{r ABEL_1}
sampleN.scABEL(CV = 0.45)
```
<small>[TOC ↩](#powertost)</small>

Iteratively adjust *α* to control the Type I Error ([Labes, Schütz](https://doi.org/10.1007/s11095-016-2006-1)). Slight heteroscedasticity (*CV*~wT~ 0.30, *CV*~wR~ 0.35), four period full replicate \"2x2x4\" study, 30 subjects, balanced sequences.
```{r ABEL_2}
scABEL.ad(CV = c(0.30, 0.35), design = "2x2x4", n = 30)
```
With the nominal *α* 0.05 the Type I Error will be inflated (0.0665). With the adjusted *α* 0.0354 (*i.e.*, a 92.92\% <span title="Confidence Interval">CI</span>) the <span title="Type I Error">TIE</span> will be controlled, although with a slight loss in power (decreases from 0.814 to 0.771).  
Consider `sampleN.scABEL.ad(CV = c(0.30, 0.35), design = "2x2x4")` to estimate the sample size which both controls the <span title="Type I Error">TIE</span> and maintains the target power. In this example 34 subjects would be required.

<small>[TOC ↩](#powertost)</small>

<span title="Average Bioequivalence with Expanded Limits">ABEL</span> cannot be applied for *AUC* (except for the <span title="World Health Organization">WHO<span>). Hence, in many cases <span title="Average Bioequivalence">ABE</span> drives the sample size. Three period full replicate \"2x2x3\" study (TRT|RTR *or* TRR|RTT).
```{r ABEL_3}
PK  <- c("Cmax", "AUC")
CV  <- c(0.45, 0.30)
# extract sample sizes and power
r1  <- sampleN.scABEL(CV = CV[1], theta0 = 0.90, design = "2x2x3",
                      print = FALSE, details = FALSE)[8:9]
r2  <- sampleN.TOST(CV = CV[2], theta0 = 0.90, design = "2x2x3",
                    print = FALSE, details = FALSE)[7:8]
n   <- as.numeric(c(r1[1], r2[1]))
pwr <- signif(as.numeric(c(r1[2], r2[2])), 5)
# compile results
res <- data.frame(PK = PK, method = c("ABEL", "ABE"), n = n, power = pwr)
print(res, row.names = FALSE)
```

<small>[TOC ↩](#powertost)</small>

#### RSABE
#### HVD(P)s
Sample size for a four period full replicate \"2x2x4\" study (any of TRTR|RTRT, TRRT|RTTR, TTRR|RRTT) assuming heteroscedasticity (*CV*~wT~ 0.40, *CV*~wR~ 0.50). Details of the sample size search suppressed.
```{r RSABE}
sampleN.RSABE(CV = c(0.40, 0.50), design = "2x2x4", details = FALSE)
```
<small>[TOC ↩](#powertost)</small>

#### NTIDs
Sample size assuming heteroscedasticity (*CV*~w~ 0.10, σ<sup>2</sup> ratio 2.5, *i.e.*, the test treatment has a substantially higher variability than the reference). TRTR|RTRT according to the [FDA’s guidance](https://www.accessdata.fda.gov/drugsatfda_docs/psg/Warfarin_Sodium_tab_09218_RC12-12.pdf). Assess additionally which one of the three components (scaled <span title="Average Bioequivalence">ABE</span>, conventional <span title="Average Bioequivalence">ABE</span>, *s*~wT~/*s*~wR~ ratio) drives the sample size.
```{r NTID2}
CV <- signif(CVp2CV(CV = 0.10, ratio = 2.5), 4)
n  <- sampleN.NTIDFDA(CV = CV)[["Sample size"]]
suppressMessages(power.NTIDFDA(CV = CV, n = n, details = TRUE))
```
The *s*~wT~/*s*~wR~ component shows the lowest probability to demonstrate <span title="Bioequivalence">BE</span> and hence, drives the sample size.

<small>[TOC ↩](#powertost)</small>

Compare that with homoscedasticity (*CV*~wT~ = *CV*~wR~ = 0.10):
```{r NTID3}
CV <- 0.10
n  <- sampleN.NTIDFDA(CV = CV, details = FALSE)[["Sample size"]]
suppressMessages(power.NTIDFDA(CV = CV, n = n, details = TRUE))
```    
Here the scaled <span title="Average Bioequivalence">ABE</span> component shows the lowest probability to demonstrate <span title="Bioequivalence">BE</span> and drives the sample size – which is much lower than in the previous example.

<small>[TOC ↩](#powertost)</small>

Comparison with *fixed* narrower limits applicable in other jurisdictions. Note that a replicate design is not required, reducing the chance of dropouts.
```{r NTID4}
CV  <- 0.10
# extract sample sizes and power
r1  <- sampleN.NTIDFDA(CV = CV, print = FALSE, details = FALSE)[8:9]
r2  <- sampleN.TOST(CV = CV, theta0 = 0.975, theta1 = 0.90,
                    design = "2x2x4", print = FALSE, details = FALSE)[7:8]
r3  <- sampleN.TOST(CV = CV, theta0 = 0.975, theta1 = 0.90,
                    design = "2x2x3", print = FALSE, details = FALSE)[7:8]
r4  <- sampleN.TOST(CV = CV, theta0 = 0.975, theta1 = 0.90,
                    print = FALSE, details = FALSE)[7:8]
n   <- as.numeric(c(r1[1], r2[1], r3[1], r4[1]))
pwr <- signif(as.numeric(c(r1[2], r2[2], r3[2], r4[2])), 5)
# compile results
res <- data.frame(method = c("FDA scaled", rep ("fixed narrow", 3)),
                  design = c(rep("2x2x4", 2), "2x2x3", "2x2x2"),
                  n = n, power = pwr, a = n * c(4, 4, 3, 2))
names(res)[5] <- "adm. #"
print(res, row.names = FALSE)
```  

<small>[TOC ↩](#powertost)</small>

### Dose-Proportionality
*CV* 0.20 (20\%), doses 1, 2, and 8 units, assumed slope *β*~0~ 1, target power 0.90.
```{r DP}
sampleN.dp(CV = 0.20, doses = c(1, 2, 8), beta0 = 1, targetpower = 0.90)
```
Note that the acceptance range of the slope depends on the ratio of the highest and lowest doses (*i.e.*, it gets tighter for wider dose ranges and therefore, higher sample sizes will be required).  
In an exploratory setting wider equivalence margins {*θ*~1~, *θ*~2~} (0.50, 2.00) were [proposed](https://doi.org/10.1002/pst.326), translating in this example to an acceptance range of `0.66667 ... 1.3333` and a sample size of only six subjects.

<small>[TOC ↩](#powertost)</small>

### Power Analysis
Explore impact of deviations from assumptions (higher *CV*, higher deviation of *θ*~0~ from 1, dropouts) on power. Assumed within-subject *CV* 0.20 (20\%), target power 0.90. Plot suppressed.
```{r PA}
res <- pa.ABE(CV = 0.20, targetpower = 0.90)
print(res, plotit = FALSE)
```
If the study starts with 26 subjects (power \~0.92), the *CV* can increase to \~0.27 **or** *θ*~0~ decrease to \~0.90 **or** the sample size decrease to 10 whilst power will still be ≥0.70.    
However, this is **not** a substitute for the “Sensitivity Analysis” recommended in [ICH-E9](https://www.ich.org/fileadmin/Public_Web_Site/ICH_Products/Guidelines/Efficacy/E9/Step4/E9_Guideline.pdf), since in a real study a combination of all effects occurs simultaneously. It is up to *you* to decide on reasonable combinations and analyze their respective power.

<small>[TOC ↩](#powertost)</small>

### Speed Comparisons
Performed on a Xeon E3-1245v3 3.4 GHz, 8 MB cache, 16 GB RAM, R `r getRversion()` 64 bit on Windows 7.

#### ABE
\"2x2\" crossover design, *CV* 0.17. Sample sizes and achieved power for the supported methods (the 1<sup>st</sup> one is the default).
```
#     method  n   power time (s)
#      owenq 14 0.80568  0.00128
#        mvt 14 0.80569  0.11778
# noncentral 14 0.80568  0.00100
#    shifted 16 0.85230  0.00096
```
The 2<sup>nd</sup> exact method is substantially slower than the 1<sup>st</sup>. The approximation based on the noncentral *t*-distribution is slightly faster but matches the 1<sup>st</sup> exact method closely. Though the approximation based on the shifted central *t*-distribution is the fastest, it *might* estimate a larger than necessary sample size. Hence, it should be used only for comparative purposes.

#### ABEL
Four period full replicate study, homogenicity (*CV*~wT~ = *CV*~wR~ 0.45). Sample sizes and achieved power for the supported methods.
```
#              function              method  n   power time (s)
#        sampleN.scABEL    ‘key’ statistics 28 0.81116   0.1348
# sampleN.scABEL.sdsims subject simulations 28 0.81196   2.5377
```
Simulating via the ‘key’ statistics is the method of choice for speed reasons.  
However, subject simulations are recommended *if*

  - the partial replicate design (TRR|RTR|RRT) is planned **and**
  - the special case of heterogenicity *CV*~wT~ > *CV*~wR~ is expected.

<small>[TOC ↩](#powertost)</small>

## Installation
You can install the released version of PowerTOST from [CRAN](https://CRAN.R-project.org) with
```{r inst}
package <- "PowerTOST"
inst    <- package %in% installed.packages()
if (length(package[!inst]) > 0) install.packages(package[!inst])
```
… and the development version from [GitHub](https://github.com/) with
```
# install.packages("remotes")
remotes::install_github("Detlew/PowerTOST")
```
Skips installation from a github remote if the [SHA-1](https://en.wikipedia.org/wiki/SHA-1) has not changed since last install. Use `force = TRUE` to force installation.

<small>[TOC ↩](#powertost)</small>

## Session Information
Inspect this information for reproducibility. Of particular importance are the versions of R and the packages used to create this workflow. It is considered good practice to record this information with every analysis.

```{r, sessioninfo}
options(width = 80)
devtools::session_info()
```

<small>[TOC ↩](#powertost)</small>
