---
title: "PowerTOST"
subtitle: "Reference-Scaled Average Bioequivalence"
author: "Helmut Schütz"
date: "`r Sys.Date()`"
lang: "en"
output:
  rmarkdown::html_vignette:
    css: vignette.css
    toc: true
vignette: >
  %\VignetteIndexEntry{PowerTOST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup}
library(PowerTOST) # attach the library
```
# Man-pages{#mans}
Man-pages of functions used in this vignette:

  * Power
    * [power.scABEL](library/html/power.scABEL.html)
    * [power.scABEL.sdsims](library/html/power.scABEL.sdsims.html)
    * [power.RSABE2L.sdsims](library/html/power.RSABE2L.sdsims.html)
    * [power.RSABE](library/html/power.RSABE.html)
    * [power.NTIDFDA](library/html/power.NTIDFDA.html)
    * [power.HVNTID](library/html/power.HVNTID.html)
  * Sample Size
    * [sampleN.scABEL](library/html/sampleN.scABEL.html)
    * [sampleN.scABEL.sdsims](library/html/sampleN.scABEL.sdsims.html)
    * [sampleN.scABEL.ad](library/html/sampleN.scABEL.ad.html)
    * [sampleN.RSABE2L.sdsims](library/html/sampleN.RSABE2L.sdsims.html)
    * [sampleN.RSABE](library/html/power.RSABE.html)
    * [sampleN.NTIDFDA](library/html/sampleN.NTIDFDA.html)
    * [sampleN.HVNTID](library/html/sampleN.HVNTID.html)
  * Helpers
    * [scABEL](library/html/scABEL.html)
    * [scABEL.ad](library/html/scABEL.ad.html)
    * [reg_const](library/html/reg_const.html)
    * [U2CVwR](library/html/U2CVwR.html)

<small>[TOC ↩](#TOC)</small>

# Defaults{#defaults}

| Parameter | Argument | Purpose | Default |
|-|----|--------------|-------|
| $\alpha$ | `alpha` | Nominal level of the test | 0.05 |
| $\pi$ | `targetpower` | <span title="typically 0.80 – 0.90">Minimum desired power</span> | 0.80 |
| $\theta_{0}$ | `theta0` | ‘True’ or assumed T/R ratio | 0.90 |
| $\theta_{1}$ | `theta1` | Lower BE limit | 0.80 |
| $\theta_{2}$ | `theta2` | Upper BE limit | 1.25 |
| *CV* | `CV` | CV | none |
| design | `design` | Planned replicate design | <span title="partial replicate">\"2x3x3\"</span> |
| regulator | `regulator` | blahblah | \"EMA\" |
| nsims | `nsims` | Number of simulations | 100,000 |
| nstart | `nstart` | Start if a previous run failed | none |
| imax | `imax` | Maximum number of iterations | 100 |
| print | `print` | Show information in the console? | TRUE |
| details | `details` | Show details of the sample size search? | FALSE |
| setseed | `setseed` | Issue a fixed seed of the random number generator? | TRUE |

Arguments have to be given as ratios, not percent.\
The *CV* is the within-subject coefficient of variation. If one value is given, homoscedasticity (equal variances) is assumed and therefore, *CV~wT~* = *CV~wR~*. If two values are given (*i.e.*, `CV=c(x, y)`) heteroscedasticity (unequal variances) is assumed, and the first value has to be *CV~wT~* and the second *CV~wR~*.

**Implemented designs**
```{r, echo=FALSE}
cols <- c(2, 9, 3)
print(known.designs()[7:12, cols], row.names = FALSE)
```
With `foo(..., details = FALSE, print = FALSE)` results are given as a data.frame with eleven columns `Design`, `alpha`,  `CVwT`, `CVwR`, `theta0`, `theta1`, `theta2`, `Sample size`, `Achieved power`, `Target power`, and `nlast`. To access *e.g.*, the sample size use either `foo[8]` or `foo[["Sample size"]]`. We suggest to use the latter in your code for clarity.

The estimated sample size gives always the *total* number of subjects (not subject/sequence).

Note that regulatory conditions are different.
```{r, echo=FALSE}
res       <- data.frame(regulator = c("EMA", "HC", "FDA"),
                        CVswitch = 0.30,  CVcap = NA, r_const = NA,
                        pe_constr = NA, method = NA,
                        stringsAsFactors = FALSE)
x         <- unlist(reg_const(regulator = "EMA"))
res[1, ]  <- x[c(1:2, 4, 3, 5:6)]
x         <- unlist(reg_const(regulator = "HC"))
res[2, ]  <- x[c(1:2, 4, 3, 6:5)]
x         <- unlist(reg_const(regulator = "FDA"))
res[3, ]  <- x[c(1:2, 4, 3, 6:5)]
res[3, 3] <- "none"
res[, 4]  <- signif(as.numeric(res[, 4]), 5)
print(res, row.names = FALSE)
```

<small>[TOC ↩](#TOC)</small>

# Sample size{#splsze}
## Example 1{#expl1}

Estimate the sample size for assumed intra-subject *CV* 0.40 (*CV~wT~* = *CV~wR~*).

### EMA ###

Average Bioequivalence with Expanding Limits (ABEL).
```{r}
sampleN.scABEL(CV = 0.55)
```
<small>[TOC ↩](#TOC)</small>

### Health Canada ###

Average Bioequivalence with Expanding Limits (ABEL).
```{r}
sampleN.scABEL(CV = 0.55, regulator = "HC")
```
<small>[TOC ↩](#TOC)</small>

### FDA ###

Reference-scaled Average Bioequivalence (RSABE).
```{r}
sampleN.RSABE(CV = 0.55)
```

<small>[TOC ↩](#TOC)</small>

# Authors{#authors}

