---
title: "Main Vignette"
lang: "en"
output:
  rmarkdown::html_vignette:
    css: vignette.css
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PowerTOST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```
| <span style="color:#006400" title="You are here">Main Vignette</span> | [ABE](ABE.html) | [RSABE](RSABE.html) | [Non-Inferiority](NI.html) | [Dose-Proportionality](DP.html) | [Power Analysis](PA.html) |
|:-:|:-:|:-:|:-:|:-:|:-:|
<div class="top"><a class="toplink" href="#" title="▲ top"> </a></div>

# Introduction
The package contains functions to calculate power and estimate sample size for various study designs used in (not only bio-) equivalence studies. Power and sample size can be obtained based on different methods, amongst them prominently the TOST procedure (two one-sided *t*-tests).\
Version `r packageVersion("PowerTOST")` built `r packageDate("PowerTOST")` with R `r getRversion()`.

For an overview of supported designs, methods, and defaults together with some basic examples see

  * [README](https://cran.r-project.org/web/packages/PowerTOST/readme/README.html) of the released version on CRAN;
  * [README](https://github.com/Detlew/PowerTOST/blob/master/README.md) of the development version on GitHub.

# Installation
You can install the released version of PowerTOST from [CRAN](https://CRAN.R-project.org) with …
```{r}
package <- "PowerTOST"
inst    <- package %in% installed.packages()
if (length(package[!inst]) > 0) install.packages(package[!inst])
```
… and the development version from [GitHub](https://github.com/) with
```
# install.packages("devtools")
devtools::install_github("Detlew/PowerTOST")
```
# Contributors
- Detlew Labes (author, maintainer)
- Helmut Schütz (author)
- Benjamin Lang (author)

# Session Information
Inspect this information for reproducibility. Of particular importance are the versions of R and the packages used to create this workflow. It is considered good practice to record this information with every analysis.
```{r, sessioninfo}
options(width = 80)
devtools::session_info()
```
